Survival in chessland

### 2018-12-16 chess/papers/survival.tex
Fork rungames for 'chessreduce', and refactor some shared code.

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their black
counterparts. The \pawn e is vastly more dead than \Pawn e. Note
somewhat satisfyingly that the \pawn c has the highest variance; this
was the most controversial among the drinking buddies
(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed pawn
in the popular Queen's Gambit (1. d4 d5 2. c4); this may be a good
example of a piece that has very different survival rates in different
opening preferences. Since the variance is otherwise very low, I did
not attempt to

@@

The fate of each piece is to either survive or die, and it does so on
one of the 64 squares. With the same replay of the 500 million games
I also kept statistics on the fates of each piece. If being a chess
piece to the death, and possessing some influence over the

Other than the bishops---which have no legal way to reach half of the
squares---every piece reaches every square in at least a thousand
games. So we have enough samples to have reasonable confidence in our
statistics, even for the most unlikely odysseys. The least mobile
pieces are the pawns, who can technically reach any square by
promoting, but are usually confined to cones emanating from their
start squares. The overall rarest fate is for \pawn f2 to die on the
a7 square, which only happened 1,244 times (however, it survived on
this square 31,438 times). This square is actually reachable without
promoting, but it would need to capture 5 times in order to get
there, which seems quite unlikely! There may even be a hidden
achievement for reaching this square this way!
% TODO: Perhaps could find a game where this happened?!
Aside from pawns, the weirdest fate is for \knight g to die on h1,
which happened 47,307 times. Corners are of course fairly garbage for
the knight, although it is twice as likely to survive on this square.

@@

That said, there are characteristic patterns for each of the pieces,
which make sense given their starting positions and patterns of
movement. You could probably guess the piece just by looking at one of
the heat maps below, although---spoiler alert---the piece is just
listed right there and they are in order. In each map, the background
color gives the

@@

\chessmap{piece0}{\Rook a8}
\chessmap{piece1}{\Knight b8}
\chessmap{piece2}{\Bishop c8}
\chessmap{piece3}{\Queen}
\chessmap{piece4}{\King}
\chessmap{piece5}{\Bishop f8}
\chessmap{piece6}{\Knight g8}
\chessmap{piece7}{\Rook h8}

@@

\chessmap{piece24}{\rook a1}
\chessmap{piece25}{\knight b1}
\chessmap{piece26}{\bishop c1}
\chessmap{piece27}{\queen}
\chessmap{piece28}{\king}
\chessmap{piece29}{\bishop f1}
\chessmap{piece30}{\knight g1}
\chessmap{piece31}{\rook h1}

>>>>

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their
Schwarzdoppelgangers. The \pawn e is vastly more dead than \Pawn e.
Note somewhat satisfyingly that the \pawn c has the highest variance;
this was the most controversial among the drinking buddies
(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed
pawn in the popular Queen's Gambit (1. d4 d5 2. c4); this may be a
good example of a piece that has very different survival rates in
different opening preferences. Since the variance is otherwise
extremely low, I only report means for the remainder of the paper.

@@

The fate of each piece is to either survive or die, and it does so on
one of the 64 squares. With the same replay of the 500 million games
I also kept statistics on the fates of each piece. If being a chess
piece to the death, and possessing some influence over where your piece
moves, it may be helpful to know where to go. Even without influence,
such knowledge could help calibrate your anxiety.

Other than the bishops---which have no legal way to reach half of the
squares---every piece ends on every square in at least a thousand
games. So we have enough samples to have reasonable confidence in our
statistics, even for the most unlikely odysseys. The least mobile
pieces are the pawns, who can technically reach any square by
promoting, but are usually confined to cones emanating from their
start squares. The overall rarest fate is for \pawn f2 to die on the
a7 square, which only happened 1,244 times (however, it survived on
this square 31,438 times). This square is actually reachable without
promoting, but it would need to capture 5 times in order to get
there, which seems quite unlikely! There may even be a hidden
achievement for reaching this square this way!
% TODO: Perhaps could find a game where this happened?!
Aside from pawns, the weirdest fate is for \knight g to die on h1,
which happened 47,307 times. Corners are of course fairly garbage for
the knight, although it is twice as likely to survive on this square.

@@

That said, there are characteristic patterns for each of the pieces,
which make sense given their starting positions and patterns of
movement. You could probably guess the piece just by looking at one of
the heat maps below, although---spoiler alert---the piece is just
listed right there and they are in order. Two independent things are
communicated in these graphics: The chance that the piece ends the
game on some square (alive or dead), and its survival chances there.
In each map, a darker background color indicates that the piece ends
on that square more often. The shade is based on the rank (64/64 black
is most common, 63/64 black is next most common, etc.) rather than the
absolute probability, since otherwise the graphic looks boring. The
number on the square is the percentage survival of the piece when it
is last seen (either being captured or surviving to the end of the
game) on that square. The four squares with the highest survival rates
are underlined for your convenience.

@@

\chessmap{piece0}{\Rook a8}
\chessmap{piece1}{\Knight b8}
\chessmap{piece2}{\Bishop c8}
\chessmap{piece3}{\Queen d8}
\chessmap{piece4}{\King e8}
\chessmap{piece5}{\Bishop f8}
\chessmap{piece6}{\Knight g8}
\chessmap{piece7}{\Rook h8}

@@

\chessmap{piece24}{\rook a1}
\chessmap{piece25}{\knight b1}
\chessmap{piece26}{\bishop c1}
\chessmap{piece27}{\queen d1}
\chessmap{piece28}{\king e1}
\chessmap{piece29}{\bishop f1}
\chessmap{piece30}{\knight g1}
\chessmap{piece31}{\rook h1}
